---
title: "Continuous Probability and the Order of Integration"
author: "Jeff Jacobs"
institute: "<a href='mailto:jj1088@georgetown.edu' target='_blank'>`jj1088@georgetown.edu`</a>"
weeknum: 6
assignment: "Lab 5"
format:
  html:
    df-print: kable
    code-fold: show
    cache: true
categories:
  - "Extra Writeups"
sidebar: mainnav
---

::: {.hidden}

```{r}
#| label: source-r-globals
source("../../dsan-globals/_globals.r")
```

:::

Problem 2e on your Lab 5 Assignment</a>, which asks you to:

::: {.callout-note title="Lab 5 Problem 2e"}

Find the probability $\Pr(X \leq Y)$

:::

is trickier than most of the other parts/problems, in the sense that it really requires us to think carefully about what exactly we're doing when we take integrals to obtain **probabilities** from **probability density functions**. So, in this writeup, I want to walk through that problem step-by-step, showing where the "trickiness" enters into the problem.

## Using Our Intuition (Without Geometry)

In this part I'm going to show how, if we ignore some of the details of the problem, we might think of the two following ways to go about obtaining the answer, which on the surface seem like equally "good" ways to go about it:

* **Approach 1: Inner integral over $Y$, outer integral over $X$**
    (1) Integrate the joint pdf $f_{X,Y}(x,y)$ from $y = x$ to the maximum value that $y$ can take on, that is, from $y = x$ to $y = 3$, then
    (2) Integrate the result from (1) over the full range of values $x$ can take on, that is, from $x = 0$ to $x = 2$.
* **Approach 2: Inner integral over $X$, outer integral over $Y$**
    (1) Integrate the joint pdf $f_{X,Y}(x,y)$ from $x = 0$ to $x = y$, then
    (2) Integrate the result from (1) over the full range of values $y$ can take on, that is, from $y = 0$ to $y = 3$

Using the joint pdf $f_{X,Y}(x,y)$ that we obtained in earlier parts of the problem,

$$
f_{X,Y}(x,y) = 3C(x^2 + 2xy^2) = \frac{1}{44}(x^2 + 2xy^2),
$$

carrying out the math for **Approach 1** gives us the following result (skipping many many steps):

$$
\begin{align*}
\Pr(X \leq Y) &= \int_{0}^{2}\int_{x}^{3}f_{X,Y}(x,y)~dy~dx \\
&= \frac{1}{132} \int_0^2 (9x^2 + 54x - 3x^3 - 2x^4)~dx \\
&= \frac{1}{132}\cdot \frac{536}{5} \approx 0.812~✅
\end{align*}
$$

So far, we can at least be a bit assured by the fact that we've obtained a value between 0 and 1, hence a valid **probability value**^[Note that this **does not mean we are necessarily *correct*!** It just means that our result doesn't "break" any laws of probability.].

Carrying out the math for **Approach 2** in the same way, however, gives us something strange:

$$
\begin{align*}
\Pr(X \leq Y) &= \int_{0}^{3}\int_{0}^{y}f_{X,Y}(x,y)~dx~dy \\
&= \frac{1}{132}\int_{0}^{3} y^3 + 3y^4 \\
&= \frac{1}{132}\cdot \frac{3321}{20} \approx 1.258~😳
\end{align*}
$$

So what happened?

## Finding The Issue By Thinking Geometrically

There are a lot of ways we could stop and try to diagnose what's going on here, but the approach that helped me most in thinking through this problem was to **plot out the space** over which we're integrating the joint pdf, **geometrically**:

Here I create what is sometimes called a "mesh grid", by running the `expand_grid()` function (from `tidyverse`, which **replaces** the base-R `expand.grid()` function you have already seen, but isn't very different) to create a grid of points, where these points' $x$ values are drawn from `x` and their $y$ values are drawn from `y`:

```{r}
#| label: create-grid
#| warning: false
library(tidyverse)
x <- seq(from = 0, to = 2, by = 0.01)
y <- seq(from = 0, to = 3, by = 0.01)
xy_df <- expand_grid(x, y)
# Display the first and last 6 rows
xy_df |> head(); xy_df |> tail()
```

Next I use the `mutate()` function to create a new variable, `x_leq_y`, on the basis of this grid, and look at the count of the number of **points** in my grid which do and do not satisfy the boolean expression `x <= y`:

```{r}
#| label: dummy-var
xy_df <- xy_df |> mutate(
  x_leq_y = as.numeric(x <= y)
)
xy_df |> group_by(x_leq_y) |> count(.drop=FALSE)
```

Finally, I use `ggplot` with the `geom_tile()` geometry to create a simple but helpful visualization of the **subset** of $[0,2] \times [0,3]$ within which the boolean condition is **true** (and where I have explicitly drawn lines showing the **boundaries** within which the point $(x,y)$ has a non-zero probability density, as well as the function $f(x) = x$, which will both come in handy later):

```{r}
#| label: plot-factor
library(latex2exp)
ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  stat_function(fun = function(x) x) +
  coord_fixed() +
  dsan_theme() +
  labs(
    fill = TeX("$X \\leq Y$"),
    title = "The Integration Space, in 2D"
  ) +
  scale_fill_discrete(labels=c("False","True"))
```

And now let's use this plot to think about what we're doing in both **Approach 1** and **Approach 2**.

### Approach 1

To think through the double integral in this case, visually, I like to imagine "sweeping" over the space across one dimension when computing the **inner** integral, then "sweeping" over the space across the other dimension when computing the **outer** integral.

Remember that the double integral in **Approach 1** was set up as follows:

$$
\begin{align*}
\Pr(X \leq Y) &= \int_{0}^{2}\int_{x}^{3}f_{X,Y}(x,y)~dy~dx
\end{align*}
$$

So, in this case, we can add some arrows to our plot to show the "sweeping" which is occurring when we compute the inner integral. Since the **inner integral** is from $y = x$ to $y = 3$, I think of arrows sweeping upwards from the line $y = x$ to the (horizontal) line $y = 3$.

So, to make the diagram on the left side of the figure below, which represents the **inner** integral, I use:

* **Dashed yellow lines** to represent the **bounds** between which we're "sweeping", and
* **Dashed arrows** to represent the **direction** of the sweeping

Then to make the diagram on the **right**, which represents the **outer** integral, I use:

* **Solid yellow lines** to represent the **bounds** between which we're sweeping, and
* **Solid arrows** to represent the **direction** of the sweeping

Which means that ultimately, after computing the two integrals, we have integrated over the **region** of the $xy$-plane bounded by the dashed and solid yellow lines.

```{r}
#| label: plot-a1-inner
#| code-fold: true
library(patchwork)
library(latex2exp)
# blw = Boundary line width
blw <- 1
# blc = Boundary line color
blc <- 'yellow'
a1_inner_xvals <- c(0.05, 0.5, 1, 1.5, 1.95)
a1_inner_df <- tibble(x=a1_inner_xvals)
a1_inner_df <- a1_inner_df |> mutate(
    xend = x,
    y = x,
    yend = 3
)
a1_inner_plot <- ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  stat_function(fun = function(x) x) +
  geom_segment(data=a1_inner_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='dashed', arrow = arrow(length = unit(0.5, "cm"))) +
  # Inner integral, lower bound
  geom_segment(x=0, y=0, xend=2, yend=2, color=blc, linewidth=blw, linetype='dashed') +
  # Inner integral, upper bound
  geom_segment(x=0, y=3, xend=2, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  coord_fixed() +
  dsan_theme() +
  labs(
    title = "A1.1: Inner Integral"
  ) +
  remove_legend()

a1_outer_yvals <- c(0.05, 0.5, 1, 1.5, 2, 2.5, 2.95)
a1_outer_df <- tibble(y=a1_outer_yvals)
a1_outer_df <- a1_outer_df |> mutate(
    yend = y,
    x = 0,
    xend = 2
)

a1_outer_plot <- ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(data=a1_inner_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='dashed', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(data=a1_outer_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='solid', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  stat_function(fun = function(x) x) +
  # Inner integral, lower bound
  geom_segment(x=0, y=0, xend=2, yend=2, color=blc, linewidth=blw, linetype='dashed') +
  # Inner integral, upper bound
  geom_segment(x=0, y=3, xend=2, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  # Outer integral, lower bound
  geom_segment(x=0, y=0, xend=0, yend=3, color=blc, linewidth=blw, linetype='solid') +
  # Outer integral, upper bound
  geom_segment(x=2, y=0, xend=2, yend=3, color=blc, linewidth=blw, linetype='solid') +
  coord_fixed() +
  dsan_theme() +
  labs(
    title = "A1.2: Outer Integral"
  ) +
  remove_legend()
a1_inner_plot + a1_outer_plot
```

Note how, since the **inner integral** in this case already **excluded** all points lying beneath the line $y = x$ (the dashed yellow diagonal line), it actually didn't matter that our **outer** integral went over this "boundary line": the overall result of the double-integration only sums up the values of the pdf in the range **bounded by the yellow lines**. In other words, even though the outer integral does "sweep" over the dashed yellow diagonal $y = x$ line, thus sweeping outside of the range of valid values of $x$, it did not have any density here to sum up, since all of the probability density in this red triangle had been left out of the inner integral.

*(If that last paragraph didn't make sense to you, don't worry about it! Just stick to the pictures for intuition)*

### Approach 2

On the other hand, recall that the double integral in **Approach 2** was set up as follows:

$$
\begin{align*}
\Pr(X \leq Y) &= \int_{0}^{3}\int_{0}^{y}f_{X,Y}(x,y)~dx~dy
\end{align*}
$$

Given **this** double integral setup, we can employ this same way of thinking about integration (as "sweeping" across a dimension) to obtain the following diagram, which helps illustrate why we obtain a value **greater than 1** in this case:

```{r}
#| label: plot-a2-inner
#| code-fold: true
a2_inner_yvals <- c(0.5, 1, 1.5, 2, 2.5, 2.95)
a2_inner_df <- tibble(y=a2_inner_yvals)
a2_inner_df <- a2_inner_df |> mutate(
    yend = y,
    x = 0.05,
    xend = y
)
ribbon_xvals <- seq(from = 2, to = 3, by = 0.01)
ribbon_df <- tibble(x=ribbon_xvals, y=ribbon_xvals)
a2_inner_plot <- ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(data=a2_inner_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='dashed', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_ribbon(data=ribbon_df, aes(x=x, ymin=y, ymax=3), alpha=0.333) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  stat_function(fun=function(x) x) +
  # Inner integral, lower bound
  geom_segment(x=0, y=0, xend=0, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  # Inner integral, upper bound
  geom_segment(x=0, y=0, xend=3, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  coord_fixed() +
  dsan_theme() +
  labs(
    title = "A2.1: Inner Integral"
  ) +
  remove_legend()

a2_outer_xvals <- c(0.05, 0.5, 1, 1.5, 1.95, 2.5, 2.95)
a2_outer_df <- tibble(x=a2_outer_xvals)
a2_outer_df <- a2_outer_df |> mutate(
    xend = x,
    y = 0,
    yend = 2.95
)

a2_outer_plot <- ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(data=a2_inner_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='dashed', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(data=a2_outer_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='solid', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_ribbon(data=ribbon_df, aes(x=x, ymin=y, ymax=3), alpha=0.333) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  stat_function(fun=function(x) x) +
  # Inner integral, lower bound
  geom_segment(x=0, y=0, xend=0, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  # Inner integral, upper bound
  geom_segment(x=0, y=0, xend=3, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  # Outer integral, lower bound
  geom_segment(x=0, y=0, xend=3, yend=0, color=blc, linewidth=blw, linetype='solid') +
  # Outer integral, upper bound
  geom_segment(x=0, y=3, xend=3, yend=3, color=blc, linewidth=blw, linetype='solid') +
  coord_fixed() +
  dsan_theme() +
  labs(
    title = "A2.2: Outer Integral"
  ) +
  remove_legend()
a2_inner_plot + a2_outer_plot
```

So we see that, whereas in **Approach 1** everything worked out since the line $y = x$ was suitable as a mathematical representation of the entire lower bound of $y$, here in **Approach 2** we **cannot** use the line $y = x$ on its own to characterize the **upper** bound on $x$, since this will include the grey triangle in the diagram above as part of the region over which we're integrating the joint pdf.

## Fixing The Issue

### A Technically-Correct Fix

Thinking through the above statement a bit further: if we want to "integrate out" the $x$ variable, the upper bound on $x$ should instead be $\min(2, y)$. In other words, by looking at the above diagram, we can see that the following **modification** to our integral in Approach 2 would technically "work" to fix the problem, if it was easy to compute:

$$
\begin{align*}
\Pr(X \leq Y) &= \int_{0}^{3}\int_{0}^{\min(2, y)}f_{X,Y}(x,y)~dx~dy
\end{align*}
$$

One way of thinking about this is, if we had set our upper bounds in **Approach 2** this way, the diagram showing the "sweeping" over the $x$ dimension would instead look like the following, where the yellow line traces out $x = \min(2,y)$ instead of $x = y$ (the black line):

```{r}
#| label: plot-a2-fixed
#| code-fold: true
a2_inner_df_fixed <- a2_inner_df |> mutate(
  xend = ifelse(xend >= 2, 2, xend)
)
a2_fixed_inner_plot <- ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(data=a2_inner_df_fixed, aes(x=x, y=y, xend=xend, yend=yend), linetype='dashed', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_ribbon(data=ribbon_df, aes(x=x, ymin=y, ymax=3), alpha=0.333) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  stat_function(fun=function(x) x) +
  # Inner integral, lower bound
  geom_segment(x=0, y=0, xend=0, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  # Inner integral, upper bound, segment 1
  geom_segment(x=2, y=2, xend=2, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  # Inner integral, upper bound, segment 2
  geom_segment(x=0, y=0, xend=2, yend=2, color=blc, linewidth=blw, linetype='dashed') +
  coord_fixed() +
  dsan_theme() +
  labs(
    title = "A2.1: Inner Integral"
  ) +
  remove_legend()

a2_outer_df_fixed <- a2_outer_df |> filter(x <= 2)

a2_fixed_outer_plot <- ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(data=a2_inner_df_fixed, aes(x=x, y=y, xend=xend, yend=yend), linetype='dashed', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(data=a2_outer_df_fixed, aes(x=x, y=y, xend=xend, yend=yend), linetype='solid', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_ribbon(data=ribbon_df, aes(x=x, ymin=y, ymax=3), alpha=0.333) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  stat_function(fun=function(x) x) +
  # Inner integral, lower bound
  geom_segment(x=0, y=0, xend=0, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  # Inner integral, upper bound, segment 1
  geom_segment(x=2, y=2, xend=2, yend=3, color=blc, linewidth=blw, linetype='dashed') +
  # Inner integral, upper bound, segment 2
  geom_segment(x=0, y=0, xend=2, yend=2, color=blc, linewidth=blw, linetype='dashed') +
  # Outer integral, lower bound
  geom_segment(x=0, y=0, xend=2, yend=0, color=blc, linewidth=blw, linetype='solid') +
  # Outer integral, upper bound
  geom_segment(x=0, y=3, xend=2, yend=3, color=blc, linewidth=blw, linetype='solid') +
  coord_fixed() +
  dsan_theme() +
  labs(
    title = "A2.2: Outer Integral"
  ) +
  remove_legend()
a2_fixed_inner_plot + a2_fixed_outer_plot
```

### Avoiding $\min()$: A Quick Fix

Since including fancy functions like $\min()$ and/or $\max()$ in our integral bounds makes things more complicated (to me, at least), one "quick fix" way to handle this would be to just take the result of the original Approach 2 integral and **subtract out** the integral of the joint pdf **within** the grey triangle in the above diagrams.

### A More Practical/"Correct" Approach

However, the more "correct" way in a sense would be to ensure that we are **never integrating over an area that is not in the support of $(X,Y)$**. To accomplish this, we can observe that our Approach 2 integral **would have worked** and **would have integrated only over the support of $(X,Y)$** if we had broken it up as follows:

$$
\Pr(X \leq Y) = \int_{0}^{2}\int_{0}^{y}f_{X,Y}(x,y)~dx~dy + \int_{2}^{3}\int_{0}^{2}f_{X,Y}(x,y)~dx~dy
$$

Thus we'll call this way of integrating **Approach 3**, which can be visualized as follows:

```{r}
#| label: a3-integral
#| code-fold: true
a31_inner_yvals <- c(0.5, 1, 1.5, 1.95)
a31_inner_df <- tibble(y=a31_inner_yvals)
a31_inner_df <- a31_inner_df |> mutate(
    yend = y,
    x = 0.05,
    xend = y
)
a31_outer_xvals <- c(0.05, 0.5, 1, 1.5, 1.95)
a31_outer_df <- tibble(x=a31_outer_xvals)
a31_outer_df <- a31_outer_df |> mutate(
    xend = x,
    y = 0.05,
    yend = 2
)
a31_plot <- ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(data=a31_inner_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='dashed', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(data=a31_outer_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='solid', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  stat_function(fun=function(x) x) +
  # Lower bound, inner integral
  geom_segment(x=0, y=0, xend=0, yend=2, color="yellow", linewidth=1, linetype='dashed') +
  # Upper bound, inner integral
  geom_segment(x=0, y=0, xend=2, yend=2, color="yellow", linewidth=1, linetype='dashed') +
  # Lower bound, outer integral
  geom_segment(x=0, y=0, xend=2, yend=0, color="yellow", linewidth=1, linetype='solid') +
  geom_segment(x=0, y=2, xend=2, yend=2, color="yellow", linewidth=1, linetype='solid') +
  coord_fixed() +
  dsan_theme() +
  labs(
    title = "A3.1: First Integral"
  ) +
  remove_legend()

a32_inner_yvals <- c(2.05, 2.5, 2.95)
a32_inner_df <- tibble(y=a32_inner_yvals)
a32_inner_df <- a32_inner_df |> mutate(
    yend = y,
    x = 0,
    xend = 1.95
)
a32_outer_xvals <- c(0.05, 0.5, 1, 1.5, 1.95)
a32_outer_df <- tibble(x=a32_outer_xvals)
a32_outer_df <- a32_outer_df |> mutate(
    xend = x,
    y = 2,
    yend = 2.95
)

a32_plot <- ggplot(xy_df, aes(x=x, y=y)) +
  geom_tile(aes(fill=factor(x_leq_y))) +
  geom_segment(data=a32_inner_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='dashed', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(data=a32_outer_df, aes(x=x, y=y, xend=xend, yend=yend), linetype='solid', arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(x=0, y=0, xend=0, yend=3) +
  geom_segment(x=0, y=0, xend=2, yend=0) +
  geom_segment(x=0, y=3, xend=2, yend=3) +
  geom_segment(x=2, y=0, xend=2, yend=3) +
  # Lower bound of inner integral
  geom_segment(x=0, y=2, xend=0, yend=3, color="yellow", linewidth=1, linetype='dashed') +
  geom_segment(x=2, y=2, xend=2, yend=3, color="yellow", linewidth=1, linetype='dashed') +
  geom_segment(x=0, y=3, xend=2, yend=3, color="yellow", linewidth=1, linetype='solid') +
  geom_segment(x=0, y=2, xend=2, yend=2, color='yellow', linewidth=1, linetype='solid') +
  stat_function(fun=function(x) x) +
  coord_fixed() +
  dsan_theme() +
  labs(
    title = "A3.2: Second Integral"
  ) +
  remove_legend()
a31_plot + a32_plot
```

So we see that the following steps give us the correct result:

* We **first** integrate the joint pdf over the region displayed on the left: the triangle with vertices at $(0,0)$, $(0,2)$, and $(2,2)$)
* We then integrate the joint pdf **separately** over the region displayed on the right: the square with vertices at $(0,2)$, $(0,3)$, $(2,2)$, and $(2,3)$
* And then we obtain the **total density of the pdf within the admissible (green) region** by **adding** these two results together

Mathematically, this gives us (skipping many steps like we did before):

$$
\begin{align*}
\Pr(X \leq Y) &= \int_{0}^{2}\int_{0}^{y}f_{X,Y}(x,y)~dx~dy + \int_{2}^{3}\int_{0}^{2}f_{X,Y}(x,y)~dx~dy \\
&= \frac{29}{165} + \frac{7}{11} = \frac{134}{165} \approx 0.812,
\end{align*}
$$

matching the result from our original integral, performed in the opposite order.

Hopefully this way of breaking down the relationship between integration and probability can help you out: if you find yourself staring at a problem for too long, try breaking your brain out of the rut by seeing if diagrams like the above can help you reason through what the bounds of your integrals should be 😎
