---
title: "Week 8: Markov Models"
subtitle: "*DSAN 5100: Probabilistic Modeling and Statistical Computing*<br>Section 03"
author: "Jeff Jacobs"
institute: "<a href='mailto:jj1088@georgetown.edu' target='_blank'>`jj1088@georgetown.edu`</a>"
date: "Thursday, October 12, 2023"
date-format: full
lecnum: 8
categories:
  - "Class Sessions"
bibliography: "../_DSAN5100.bib"
format:
  revealjs:
    cache: false
    footer: "DSAN 5100-03 Week 8: Markov Models"
    output-file: slides.html
    df-print: kable
    code-fold: true
    html-math-method: mathjax
    scrollable: true
    slide-number: true
    section-divs: false
    simplemenu:
      flat: true
      barhtml:
        header: "<div class='menubar'><span style='position: absolute; left: 8; padding-left: 8px;'><a href='./index.html'>&larr; Return to Notes</a></span><ul class='menu'></ul></div>"
      scale: 0.5
    number-sections: false
    footnotes-hover: true
    tbl-cap-location: bottom
    theme: [default, "../_style-slides.scss"]
    revealjs-plugins:
      - simplemenu
    include-in-header:
      text: "<link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap@5.2.3/dist/css/bootstrap.min.css' integrity='sha384-rbsA2VBKQhggwzxH7pPCaAqO46MgnOM80zW1RWuH61DGLwZJEdK2Kadq2F9CUG65' crossorigin='anonymous'><link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap-icons@1.10.5/font/bootstrap-icons.css'>"
  html:
    output-file: index.html
    cache: false
---

::: {.content-visible unless-format="revealjs"}

<center>
<a class="h2" href="./slides.html" target="_blank">Open slides in new window &rarr;</a>
</center>

:::

# Schedule {.smaller .small-title .crunch-title .crunch-callout data-name="Schedule"}

::: {.hidden}

```{r}
#| label: r-source-globals
source("../_globals.r")
```

{{< include ../_globals-tex.qmd >}}

:::

Today's Planned Schedule:


| | Start | End | Topic | Recording |
|:- |:- |:- |:- |:-:|
| **Lecture** | 12:30pm | 1:00pm | <a href="#markov-chains">Markov Chains, Hidden Markov Models &rarr;</a> | <a href="../recordings/recording-w08-1.html" target="_blank"><i class="bi bi-film"></i></a> |
| | 1:00pm | 1:30pm | <a href="#pagerank">PageRank &rarr;</a> | <a href="../recordings/recording-w08-2.html" target="_blank"><i class="bi bi-film"></i></a> |
| | 1:30pm | 2:00pm | <a href="#matrix-magic">Matrix Magic &rarr;</a> | <a href="../recordings/recording-w08-3.html" target="_blank"><i class="bi bi-film"></i></a> |
| **Break!** | 2:00pm | 2:10pm | | |
| **Lab** | 2:10pm | 2:40pm | <a href="#lab-demonstrations">Lab &rarr;</a> | <a href="../recordings/recording-w08-4.html" target="_blank"><i class="bi bi-film"></i></a> |
| | 2:40pm | 3:00pm | Student Presentation | |

: {tbl-colwidths="[14,12,12,50,12]"}

# Motivation {data-stack-name="Motivation"}

## The Course So Far {.crunch-center .crunch-lists .crunch-title .crunch-ul}

* **Logic**: Reasoning about *T/F statements* using **and** and **or**
* $\rightarrow$ **Set Theory**: Reasoning about *collections of objects* and their **union** and **intersection**

<center class='smaller-text'><i class='bi bi-arrow-up pe-1'></i>Deterministic<i class='bi bi-arrow-up ps-1'></i></center>
<hr style='margin-bottom: 1px !important; margin-top: 1px !important;'>
<center class='smaller-text'><i class='bi bi-arrow-down pe-2'></i>Probabilistic<i class='bi bi-arrow-down ps-2'></i></center>

* $\rightarrow$ **Probability**: Assigning "likelihood values" in $[0,1]$ to logical/set-theoretic statements about **outcomes**
* $\rightarrow$ **Random Variables**: Doing **math** (beyond **and** and **or**) with the probabilities of these uncertain outcomes
* All have been **reasoning about uncertainty** in more and more complex ways!

## The New Dimension: Evolution Through *Time*

* **Markov models** are a stepping-stone towards full-on **time series analysis**: reasoning about uncertainty **over time**
* So why are **time series** the topic of an entire course, while **Markov models** only the topic of one week in DSAN5000?

## Recall: Definition of Conditional Probability

$$
\Pr(A \mid B) \definedas \frac{\Pr(A, B)}{\Pr(B)} \genfrac{}{}{0pt}{}{\leftarrow A\text{ and }B\text{ happen}}{\leftarrow \text{In world where }\Omega = B}
$$

## The Markov Property

$$
P(\text{future} \mid \text{present}, {\color{orange}\text{past}}) = P(\text{future} \mid \text{present})
$$

![](images/markov_property.svg){fig-align="center"}

## The Markov Assumption

Often stated in many different (confusing) ways, but think of **anterograde amnesia**:

<video width="100%" height="60%" controls>
  <source src="https://jpj.georgetown.domains/dsan5100-scratch/memento_chase.mp4" type="video/mp4">
</video>

## Keeping Track of States

* Since Markov models come more from engineering than "pure" math/probability, we have "fancier" objects we're keeping track of, rather than just Random Variables
* We now keep track of the **state** of a **system** over time: this could be a single RV, but could be a **collection** of multiple (potentially *dependent*) RVs

## What Information Should We Keep Track Of? {.smaller .title-12 .crunch-title}

* Sometimes answer seems "obvious", lulls us into false sense of confidence (that we don't have to think too hard about it)[^complexity]

| System | State Space |
| - | - |
| Chess | Position of each piece, whose turn it is |
| Indoor Robot | Model of room + objects inside it |
| Predator-Prey Ecosystem | Relative species populations |
| Move Left/Move Right Game | ? |
| Weather Prediction | ? |
| Music | ? |

* Tension between **too little** information (cannot model phenomena of interest) vs. **too much** information (solving model becomes intractable)

[^complexity]: To snap out of this lull, keep in mind that engineering/nonlinear dynamical systems programs have <a href='https://www.youtube.com/watch?v=rXnZ-HFoOz8&list=PLF0b3ThojznTzAA7bfLWh4RKzRrwNF4L0' target='_blank'>entire courses</a> **just** on state space representations

## Modeling Music: Where is the Downbeat? {.smaller .title-12 .crunch-figures .crunch-images .crunch-quarto-layout-panel .nostretch}

::: {layout-ncol=2}

{{< video https://www.youtube.com/watch?v=xe_iCkFsQKE width="100%" height="300" >}}

{{< video https://www.youtube.com/watch?v=XbU9UUwxBxA width="100%" height="300" >}}

:::

::: {.music-automata}
::: {layout-ncol=3}

![](images/automata_34.svg){fig-align="center" width="300"}

![](images/automata_44.svg){fig-align="center" width="350"}

![](images/automata_54.svg){fig-align="center" width="350"}

:::
:::

## State Space Choice $\rightarrow$ Information Loss {.smaller .nostretch .title-12}

* Recall the Move Left/Move Right Game (especially from <a href='https://jjacobs.me/dsan5100-03/463a01339cf0f456ba54a1849df50d1a22c247e3/writeups/lab-4-prep/DSAN5100_Lab_4_Prep.html' target='_blank'>Lab 4 Prep</a>)

::: columns
::: {.column width="45%"}

![](images/random_walk_wide.svg){fig-align="center"}

:::
::: {.column width="55%"}

<center>
Possible State Spaces:
</center>

| History vector | Position | Steps | #L | #R |
| - | -:|:-:|:-:|:-:|
| $()$ | $x = 0$ | 0 | 0 | 0 |
| $(L,R)$ | $x = 0$ | 2 | 1 | 1 |
| $(R,L)$ | $x = 0$ | 2 | 1 | 1 |
| $(R,L,R,$<br>$~L,R,L)$ | $x = 0$ | 6 | 3 | 3 |
| $(L)$ | $x = -1$ | 1 | 1 | 0 |
| $(L,L,R)$ | $x = -1$ | 3 | 2 | 1 |

:::
:::

# Markov Models {data-stack-name="Markov Models"}

<!-- ## Four Types {.smaller}

| | Fully Observable | Partially Observable |
| - | - | - |
| **Autonomous** | Markov Chain<br> | Hidden Markov Model |
| **Controlled** | Markov Decision Process | Partially-Observable MDP | -->

## Finite-State Automata {.nostretch .crunch-title}

(Deterministic!) Only "accepts" strings with even number of `1`s:

::: columns
::: {.column width="44%"}

![](images/automata.svg){fig-align="center" width="80%"}

:::
::: {.column width="56%"}

```{=html}
<table>
<thead>
  <tr>
  <th>Input String</th>
  <th class='split-table'>Result</th>
  <th>Input String</th>
  <th>Result</th>
  </tr>
</thead>
<tbody>
  <tr>
    <td><span data-qmd="$\varepsilon$"></span></td>
    <td class='split-table'>✅</td>
    <td><span data-qmd="`01`"></span></td>
    <td><span data-qmd="<i class='bi bi-x-lg' style='color: red;'></i>"></span></td>
  </tr>
  <tr>
    <td><span data-qmd="`0`"></span></td>
    <td class='split-table'>✅</td>
    <td><span data-qmd="`10`"></span></td>
    <td><span data-qmd="<i class='bi bi-x-lg' style='color: red;'></i>"></span></td>
  </tr>
  <tr>
    <td><span data-qmd="`1`"></span></td>
    <td class='split-table'><span data-qmd="<i class='bi bi-x-lg' style='color: red;'></i>"></span></td>
    <td><span data-qmd="`1000000`"></span></td>
    <td><span data-qmd="<i class='bi bi-x-lg' style='color: red;'></i>"></span></td>
  </tr>
  <tr>
    <td><span data-qmd="`00`"></span></td>
    <td class='split-table'>✅</td>
    <td><span data-qmd="`10000001`"></span></td>
    <td>✅</td>
  </tr>
</tbody>
</table>
```

<!-- | Input String | Result | Input String | Result |
| - | - | - | - |
| $\varepsilon$ | ✅ | `01` | <i class='bi bi-x-lg' style='color: red;'></i> |
| `0` | ✅ | `10` | <i class='bi bi-x-lg' style='color: red;'></i> |
| `1` | <i class='bi bi-x-lg' style='color: red;'></i> | `1000000` | <i class='bi bi-x-lg' style='color: red;'></i> |
| `00` | ✅ | `10000001` | ✅ | -->

:::
:::

* ...But we're trying to model **probabilistic** evolution!

## Enter Markov Chains

::: columns
::: {.column width="50%"}

<center>
Graphically
</center>

![](images/basic_no_self_loops.svg){fig-align="center" width="60%"}

![](images/basic_self_loops.svg){fig-align="center" width="60%"}

:::
::: {.column width="50%"}

<center>
Mathematically
</center>

$$
\begin{array}{c c} 
& \begin{array}{c c c} 1 & 2 & 3 \\ \end{array} \\
\begin{array}{c c c}1 \\ 2 \\ 3 \end{array} &
\left[
\begin{array}{c c c}
0 & 1/2 & 1/2 \\
1/3 & 0 & 2/3 \\
1/3 & 2/3 & 0
\end{array}
\right]
\end{array}
$$


$$
\begin{array}{c c} 
& \begin{array}{c c c} 1 & 2 & 3 \\ \end{array} \\
\begin{array}{c c c}1 \\ 2 \\ 3 \end{array} &
\left[
\begin{array}{c c c}
1/2 & 1/3 & 1/6 \\
1/10 & 1/2 & 2/5 \\
1/8 & 3/8 & 1/2
\end{array}
\right]
\end{array}
$$

:::
:::

## Hidden Markov Models {data-name="HMMs"}

* Use **observed** data to infer **unobserved** variables

![](images/hmm.svg){fig-align="center"}

* (What our brains are doing, most of the time!)

# PageRank {data-stack-name="PageRank"}

## PageRank (Matrix Magic)

* What is the relevance of this abstract topic? ...🤑

```{r}
#| label: google-graph
#| echo: true
library(readr)
library(ggplot2)
goog_df <- read_csv("assets/google_yearly_revenue.csv")
ggplot(goog_df, aes(x=year, y=revenue_billions)) +
  geom_bar(stat="identity", fill=cbPalette[1]) +
  labs(
    title = "Google Yearly Revenue, 2002-2022",
    x = "Year",
    y = "Revenue (Billion USD)"
  ) +
  dsan_theme("full")
```

* PageRank = The "spark" that ignited the Google flame

## PageRank Visualized {.smaller}

* Nodes = Webpages, Edges = Links

![](images/pagerank.svg){fig-align="center"}

* Goal: Rank the relative "importance" of a site $S_i$, **taking into account** the importance of other sites that **link to** $S_i$
  * "Important" sites are linked to often, and linked to often **by other important sites**

## Chickens and Eggs {.smaller}

* Paradoxical at first: how are we supposed to figure out the **importance of a site** $S_i$, when that's determined by
  * the **importance of sites $S_j$ that link to $S_i$**, which is determined by
    * the **importance of sites $S_k$ that link to sites $S_j$**, which is determined by
      * the **importance of the sites $S_\ell$ that link to those sites $S_k$**, which is determined by...

$$
\begin{align*}
\mathsf{Importance}(S_i) &= f(\mathsf{Importance}(S_{j \rightarrow i})) = f(f(\mathsf{Importance}(S_{k \rightarrow j \rightarrow i}))) \\
&= f(f(f(\mathsf{Importance}(S_{\ell \rightarrow k \rightarrow j \rightarrow i})))) = \cdots
\end{align*}
$$

::: {.aside}
Sanity hint: Remember infinite sums from calculus! They can converge, despite having infinitely-many terms... This is something like that, but for **recursion** (the mathematical term for an object whose definition refers to itself)
:::

## Resolving Recursive Definitions {.smaller}

* We **can** compute this importance ranking, despite its recursive definition!
* Recall, for example, the Fibonacci sequence: $1, 1, 2, 3, 5, 8, 13, 21, \ldots$
* Defined recursively!

$$
f(n) = \begin{cases}
1 & n = 1\text{ or }n = 2 \\
f(n-2) + f(n-1) & n > 2
\end{cases}
$$

* And yet, a guy named <a href="https://en.wikipedia.org/wiki/Daniel_Bernoulli" target="_blank">Bernoulli</a> figured out

$$
f(n) = \frac{\varphi^n - \psi^n}{\varphi - \psi} = \frac{\varphi^n - \psi^n}{\sqrt{5}},
$$

where $\varphi$ is the "Golden Ratio" $\frac{1 + \sqrt{5}}{2}$ and $\psi$ its conjugate $\frac{1 - \sqrt{5}}{2}$.

## The PageRank Process {.smaller .crunch-title}

* Every site starts with **equal** PageRank score: $r^{(0)}_1 = r^{(0)}_2 = r^{(0)}_3 = \frac{1}{3}$.
* Each link $S_i \rightarrow S_j$ is a **vote of confidence** that $S_i$ is giving to $S_j$
* At each time $t$, a site $S_i$ **"spends"** whatever voting power it currently has ($r^{(t)}_i$) on the sites it links to.
  * $S_1$ casts one vote for itself and one vote for $S_2$, thus spending $\frac{1}{2}$ of its total PageRank on itself and $\frac{1}{2}$ of its total PageRank on $S_2$.
* **State** of the process at time $t$: $\mathbf{r}^{(t)} = \begin{bmatrix}r^{(t)}_1 & r^{(t)}_2 & r^{(t)}_3\end{bmatrix}^\top$
* Can form a **matrix** specifying exactly how this state **evolves** from time $t$ to time $t+1$!

$$
\mathbf{E} = \begin{bmatrix}
\frac{1}{2} & \frac{1}{2} & 0 \\
\frac{1}{2} & 0 & 1 \\
0 & \frac{1}{2} & 0
\end{bmatrix} \; \leadsto \; \mathbf{r}^{(t+1)} = \mathbf{E}\mathbf{r}^{(t)}
$$

::: {.notes}
* Given the "$S_1$ casts one vote for itself..." part, can you say exactly what $S_1$ will "spend" on itself and on $S_2$ at time $t = 0$ (in the first round)?
:::

## {.smaller}

* We can use $\mathbf{E}$ to figure out the state at each step, starting from $t = 0$!

$$
\begin{array}{c@{}c@{}c@{}c@{}c@{}c@{}c}
\mathbf{r}^{(1)} & = & \mathbf{E}\mathbf{r}^{(0)} & = & \begin{bmatrix}
\frac{1}{2} & \frac{1}{2} & 0 \\
\frac{1}{2} & 0 & 1 \\
0 & \frac{1}{2} & 0
\end{bmatrix}\begin{bmatrix}
\frac{1}{3} \\
\frac{1}{3} \\
\frac{1}{3}\end{bmatrix} & = & \begin{bmatrix}
\frac{1}{3} \\
\frac{1}{2} \\
\frac{1}{6}
\end{bmatrix} \\
~ & ~ & ~ & ~ & ~ & \swarrow & ~ \\
\mathbf{r}^{(2)} & = & \mathbf{E}\mathbf{r}^{(1)} & = & \begin{bmatrix}
\frac{1}{2} & \frac{1}{2} & 0 \\
\frac{1}{2} & 0 & 1 \\
0 & \frac{1}{2} & 0
\end{bmatrix}\begin{bmatrix}\frac{1}{3} \\ \frac{1}{2} \\ \frac{1}{6}\end{bmatrix} & = & \begin{bmatrix}\frac{5}{12} \\ \frac{1}{3} \\ \frac{1}{4}\end{bmatrix} \\
~ & ~ & ~ & ~ & ~ & \swarrow & ~ \\
\mathbf{r}^{(3)} & = & \mathbf{E}\mathbf{r}^{(2)} & = & \cdots & ~ & ~
\end{array}
$$

## Matrix Magic {.smaller}

* (1) Won't we just have to run this forever? (2) How do we know it'll converge to something?
* Answers: (1) no, (2) because **Markov matrix magic**
* "Steady state" = state where $\mathbf{r}^{(t)} = \mathbf{r}^{(t+1)} = \mathbf{r}^{(t+2)} = \cdots \definedas \mathbf{r}^*$. But this means

$$
\mathbf{r}^{(t+1)} = \mathbf{r}^{(t)} \iff \mathbf{E}\mathbf{r}^{(t)} = \mathbf{r}^{(t)} \iff \mathbf{E}\mathbf{r}^* = \mathbf{r}^*
$$

* This $\mathbf{r}^*$ is---by definition!---an **Eigenvector** of $\mathbf{E}$, corresponding to **Eigenvalue** $\lambda = 1$![^lambda]

[^lambda]: This is because an **Eigenvalue-Eigenvector pair** for a matrix $\mathbf{M}$ is a vector $\mathbf{v}$ and scalar value $\lambda$ which satisfy $\mathbf{M}\mathbf{v} = \lambda \mathbf{v}$. In words: the result of (left) **matrix-multiplying** $\mathbf{v}$ by $\mathbf{M}$ is the same as **scalar-multiplying** $\mathbf{v}$ by a factor of $\lambda$. In our case the **Eigenvector** is $\mathbf{r}^*$ and the **Eigenvalue** is $\lambda = 1$, since $\mathbf{E}\mathbf{r}^* = 1 \cdot \mathbf{r}^*$.<br>For the math-curious, there are lots of fun results from matrix theory which assure us that $\mathbf{E}$ is **guaranteed** to have principal eigenvalue $\lambda = 1$ 💆

::: {.notes}
* In my opinion, along with e.g. insolubility of the quintic, this is maybe the most mind-blowing case of math magic :3
:::

## Solving the Matrix Magic {.smaller .crunch-title}

* Since we already know the **Eigenvalue** of interest, $\lambda = 1$, all that's left is solving for its corresponding **Eigenvector**:

$$
\mathbf{E}\mathbf{r}^* = \mathbf{r}^* \iff \mathbf{E}\mathbf{r}^* - \mathbf{r}^* = \mathbf{0} \iff (\mathbf{E} - \mathbf{I})\mathbf{r}^* = \mathbf{0}
$$

* Written out, we see that this gives us a system of linear equations:

$$
\begin{bmatrix}
\frac{1}{2} & \frac{1}{2} & 0 \\
\frac{1}{2} & 0 & 1 \\
0 & \frac{1}{2} & 0
\end{bmatrix}\begin{bmatrix}r^*_1 \\ r^*_2 \\ r^*_3\end{bmatrix} = \begin{bmatrix}0 \\ 0 \\ 0\end{bmatrix} \iff \begin{array}{ccccccc}\frac{1}{2}r^*_1 & + & \frac{1}{2}r^*_2 & ~ & ~ & = & 0 \\ \frac{1}{2}r^*_1 & ~ & ~ & + & r^*_3 & = & 0 \\ ~ & ~ & \frac{1}{2}r^*_2 & ~ & ~ & = & 0\end{array}
$$

which we can solve however we want!

* To handle the fact that this system is **underspecified**, we impose the additional restriction that $r^*_1 + r^*_2 + r^*_3 = 1$, so that the ranks form a **probability distribution**...

## The Point of All This {.crunch-title}

* The final restriction $r^*_1 + r^*_2 + r^*_3 = 1$ ensures that the resulting PageRank values form a **probability distribution**
* This is called the **Stationary Distribution** of the **Markov chain**: represents the probability that a **random walker** through the chain will be at page $S_i$ at a given time!
  * Equivalently: **expected proportion of total walking time** a random-walker will spend at each node
* **Every Markov chain has a Stationary Distribution**! This fact has cool implications even above and beyond the Google $$$ implications 😜
